---
title: "Rural Malawi ESBL-producing E. coli Genomics"
author: "Angus O'Ferrall"
date: "2026-06-12"
output: html_document
---

# Preamble

To run this code, first download the associated data to a single folder called 'data'.
The 'data' folder should be in your current local file path.
Also create a folder called 'figures' where the generated figures will be saved.

```{r}
# Attach packages
knitr::opts_chunk$set(echo=TRUE)
library(knitr)
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(forcats)
library(stringr)
library(ggtree)
library(ggtreeExtra)
library(ape)
library(phangorn)
library(scales)
library(ggnewscale)
library(sf)
library(ggspatial)
library(purrr)
library(patchwork)
library(ComplexHeatmap)
library(circlize)
library(grid)
library(FSA)
library(ggpubr)
library(igraph)
library(ggraph)
```

# Participant data

```{r}
# Import and tidy the participant data

# Import the data
part_data <- read.csv("data/Participant_data.csv", na.strings = "")

# Set missing data to NA
part_data[part_data == 99 ] <- NA

# Tidy up categorical variables
part_data$sex <- as.factor(part_data$sex)
part_data$follow_up <- as.factor(part_data$follow_up)
part_data$where_defecate <- as.factor(part_data$where_defecate)
part_data <- part_data %>%
  mutate(
    where_defecate = case_when(
      where_defecate %in% c("bush", "other") ~ "other",
      TRUE ~ where_defecate
    )
  )
part_data$drinking_water_source <- as.factor(part_data$drinking_water_source)
part_data$fu_piped_water_access <- as.factor(part_data$fu_piped_water_access)

# Re-name variables for colonisation in 2023 (baseline) and 2024 (follow-up)
names(part_data)[names(part_data) == "baseline_ESBL_E_coli"] <- "2023"
names(part_data)[names(part_data) == "fu_ESBL_E_coli"] <- "2024"

# Convert and label categories
part_data$district <- factor(part_data$district, levels = c(0, 1), labels = c('Nsanje', 'Mangochi'))
part_data$'2023' <- factor(part_data$'2023', levels = c(0, 1), labels = c('No', 'Yes'))
part_data$'2024' <- factor(part_data$'2024', levels = c(0, 1), labels = c('No', 'Yes'))
part_data$baseline_multihousehold_strain <- factor(part_data$baseline_multihousehold_strain, levels = c(0, 1), labels = c('No', 'Yes'))
part_data$fu_multihousehold_strain <- factor(part_data$fu_multihousehold_strain, levels = c(0, 1), labels = c('No', 'Yes'))

# Identify participants colonised with a multi-household strain at any point
part_data$ever_multihousehold_strain <- factor(
  ifelse(
    ifelse(is.na(part_data$baseline_multihousehold_strain), "No",
           as.character(part_data$baseline_multihousehold_strain)) == "Yes" |
    ifelse(is.na(part_data$fu_multihousehold_strain), "No",
           as.character(part_data$fu_multihousehold_strain)) == "Yes",
    "Yes", "No"
  ),
  levels = c("No", "Yes")
)
```

```{r}
# Age
summary(part_data$age_at_baseline)
```

```{r}
# Categorical variable summaries
categorical_vars <- c(
  "sex",
  "follow_up",
  "where_defecate",
  "drinking_water_source"
)

# Generate summary of categorical variables
lapply(categorical_vars, function(v) {
  part_data %>%
    count(.data[[v]]) %>%
    mutate(percent = round(100 * n / sum(n), 1))
})
```

## Per district

### Mangochi

```{r}
# Filter the data
part_data_mangochi <- part_data %>%
  filter(district == "Mangochi")
```

```{r}
# Age
summary(part_data_mangochi$age_at_baseline)
```

```{r}
# Generate summary of categorical variables
lapply(categorical_vars, function(v) {
  part_data_mangochi %>%
    count(.data[[v]]) %>%
    mutate(percent = round(100 * n / sum(n), 1))
})
```

### Nsanje

```{r}
# Filter the data
part_data_nsanje <- part_data %>%
  filter(district == "Nsanje")
```

```{r}
# Age
summary(part_data_nsanje$age_at_baseline)
```

```{r}
# Generate summary of categorical variables
lapply(categorical_vars, function(v) {
  part_data_nsanje %>%
    count(.data[[v]]) %>%
    mutate(percent = round(100 * n / sum(n), 1))
})
```

### Differences between locations

```{r}
# Age
wilcox.test(
  age_at_baseline ~ district,
  data = part_data
)
```

```{r}
# Sex
sex_tab <- table(
  part_data$sex,
  part_data$district
)

fisher.test(sex_tab)
```

```{r}
# Follow-up
fu_tab <- table(
  part_data$follow_up,
  part_data$district
)

fisher.test(fu_tab)
```

```{r}
# Latrine
latrine_tab <- table(
  part_data$where_defecate,
  part_data$district
)

fisher.test(latrine_tab)
```

```{r}
# Drinking water
water_tab <- table(
  part_data$drinking_water_source,
  part_data$district
)

fisher.test(water_tab)
```

## Carriage dynamics

```{r}
# Mutate data to show dynamics in participants followed-up
fu_part_data <- part_data %>%
  filter(follow_up == "yes") %>%
  mutate('change' = case_when(
    `2023` == "No" & `2024` == "No" ~ "Stayed negative",
    `2023` == "Yes" & `2024` == "Yes" ~ "Stayed positive",
    `2023` == "No" & `2024` == "Yes" ~ "Became positive",
    `2023` == "Yes" & `2024` == "No" ~ "Became negative"
  ))

# Set the levels for the change variable to the desired order
fu_part_data$change <- factor(fu_part_data$change, 
                              levels = c("Stayed negative", "Became positive", "Became negative", "Stayed positive"))

# Summarise counts of participants for each type of change, and negate values for "negative" changes
summary_data <- fu_part_data %>%
  count(change) %>%
  mutate(n = ifelse(change %in% c("Stayed negative", "Became negative"), -n, n))

# Add the 'baseline' column based on the conditions
summary_data <- summary_data %>%
  mutate(baseline = case_when(
    change %in% c('Stayed negative', 'Became positive') ~ 'Negative',
    change %in% c('Stayed positive', 'Became negative') ~ 'Positive'
  ))

# Create the plot
ESBL_dynamics <- ggplot(summary_data, aes(x = change, y = n)) +
  geom_bar(stat = "identity", aes(fill = baseline), colour = "black", linewidth = 0.3, width = 0.7) +
  scale_y_continuous(labels = abs) +
  scale_fill_manual(values = c("Negative" = "cornflowerblue", 
                                "Positive" = "firebrick")) +
  labs(fill = "Baseline status") +
  labs(x = "ESBL-Ec carriage dynamic", 
       y = "Number of participants") +
  theme_minimal() +
  theme(
    axis.title.x = element_text(margin = margin(t = 10)),  # Space above x-axis label
    axis.title.y = element_text(margin = margin(r = 10)),  # Space to the right of y-axis label
    plot.margin = margin(r = 30)  # Extra space on the right of the plot for legend
  ) +
  geom_hline(yintercept = 0, linetype = "dashed")

# Save the plot
ggsave("figures/ESBL_dynamics.png", ESBL_dynamics, height = 5, dpi = 600)

# Print the plot
print(ESBL_dynamics)
```

## Change in prevalence

```{r}
# Specify columns to look at prevalence in
cols <- c("2023", "2024", "baseline_multihousehold_strain", "fu_multihousehold_strain", "ever_multihousehold_strain")

# Calculate prevalences in 2023 and 2024 of colonisation and colonisation with a multihousehold strain
prevalence <- sapply(part_data[cols], function(x) {
  mean(x == "Yes", na.rm = TRUE)
})

print(prevalence)
```

```{r}
# Calculate prevalences in participants followed-up
prevalence <- sapply(fu_part_data[cols], function(x) {
  mean(x == "Yes", na.rm = TRUE)
})

print(prevalence)
```
```{r}
# Check for significant rise in prevalence in colonisation follow-up participants using McNemar's test for paired data
colonisation_tab <- table(fu_part_data$'2023', fu_part_data$'2024')
mcnemar.test(colonisation_tab)
```

## Prevalence differences between locations

```{r}
# Baseline prevalence
bl_prev_tab <- table(
  part_data$"2023",
  part_data$district
)

bl_prev_tab

chisq.test(bl_prev_tab, correct = FALSE)
```

```{r}
# Follow-up prevalence
fu_prev_tab <- table(
  fu_part_data$"2024",
  fu_part_data$district
)

fu_prev_tab

chisq.test(fu_prev_tab, correct = FALSE)
```

## Prevalence summaries with confidence intervals

### All participants

```{r}
# Define the variables of interest
vars <- c("2023", "2024", "ever_multihousehold_strain")

# Initialize the summary dataframe
prevalence_summary <- data.frame(
  Variable = vars,
  Positive_Count = NA,
  Total_N = NA,
  Prevalence_Percent = NA,
  CI_Lower = NA,
  CI_Upper = NA
)

# Loop through variables and calculate values
for (i in seq_along(vars)) {
  var <- vars[i]
  x <- part_data[[var]]
  pos <- sum(x %in% c("Yes"), na.rm = TRUE)
  total <- sum(!is.na(x))
  test <- prop.test(pos, total)

  prevalence_summary$Positive_Count[i] <- pos
  prevalence_summary$Total_N[i] <- total
  prevalence_summary$Prevalence_Percent[i] <- round(100 * pos / total, 1)
  prevalence_summary$CI_Lower[i] <- round(100 * test$conf.int[1], 1)
  prevalence_summary$CI_Upper[i] <- round(100 * test$conf.int[2], 1)
}

# Print the table
print(prevalence_summary)
```

### Baseline prevalence in follow-up participants

```{r}
# Define the variables of interest
vars <- c("2023")

# Initialize the summary dataframe
prevalence_summary <- data.frame(
  Variable = vars,
  Positive_Count = NA,
  Total_N = NA,
  Prevalence_Percent = NA,
  CI_Lower = NA,
  CI_Upper = NA
)

# Loop through variables and calculate values
for (i in seq_along(vars)) {
  var <- vars[i]
  x <- fu_part_data[[var]]
  pos <- sum(x %in% c("Yes"), na.rm = TRUE)
  total <- sum(!is.na(x))
  test <- prop.test(pos, total)

  prevalence_summary$Positive_Count[i] <- pos
  prevalence_summary$Total_N[i] <- total
  prevalence_summary$Prevalence_Percent[i] <- round(100 * pos / total, 1)
  prevalence_summary$CI_Lower[i] <- round(100 * test$conf.int[1], 1)
  prevalence_summary$CI_Upper[i] <- round(100 * test$conf.int[2], 1)
}

# Print the table
print(prevalence_summary)
```

### Mangochi District

```{r}
# Define the variables of interest
vars <- c("2023", "2024")

# Initialize the summary dataframe
prevalence_summary <- data.frame(
  Variable = vars,
  Positive_Count = NA,
  Total_N = NA,
  Prevalence_Percent = NA,
  CI_Lower = NA,
  CI_Upper = NA
)

# Loop through variables and calculate values
for (i in seq_along(vars)) {
  var <- vars[i]
  x <- part_data_mangochi[[var]]
  pos <- sum(x %in% c("Yes"), na.rm = TRUE)
  total <- sum(!is.na(x))
  test <- prop.test(pos, total)

  prevalence_summary$Positive_Count[i] <- pos
  prevalence_summary$Total_N[i] <- total
  prevalence_summary$Prevalence_Percent[i] <- round(100 * pos / total, 1)
  prevalence_summary$CI_Lower[i] <- round(100 * test$conf.int[1], 1)
  prevalence_summary$CI_Upper[i] <- round(100 * test$conf.int[2], 1)
}

# Print the table
print(prevalence_summary)
```

### Nsanje District

```{r}
# Define the variables of interest
vars <- c("2023", "2024")

# Initialize the summary dataframe
prevalence_summary <- data.frame(
  Variable = vars,
  Positive_Count = NA,
  Total_N = NA,
  Prevalence_Percent = NA,
  CI_Lower = NA,
  CI_Upper = NA
)

# Loop through variables and calculate values
for (i in seq_along(vars)) {
  var <- vars[i]
  x <- part_data_nsanje[[var]]
  pos <- sum(x %in% c("Yes"), na.rm = TRUE)
  total <- sum(!is.na(x))
  test <- prop.test(pos, total)

  prevalence_summary$Positive_Count[i] <- pos
  prevalence_summary$Total_N[i] <- total
  prevalence_summary$Prevalence_Percent[i] <- round(100 * pos / total, 1)
  prevalence_summary$CI_Lower[i] <- round(100 * test$conf.int[1], 1)
  prevalence_summary$CI_Upper[i] <- round(100 * test$conf.int[2], 1)
}

# Print the table
print(prevalence_summary)
```

## Associations with multi-household strain colonisation

### Anytime colonisation with a multi-household strain

```{r}
# Logistic regression model adjusting for follow-up status
fit <- glm(ever_multihousehold_strain ~ age_at_baseline + sex + district + follow_up,
  family = binomial(logit),
  data = part_data
)
summary(fit)
```

```{r}
# Get exponentiated model coefficients (adjusted odds ratios) and 95% CIs
exp(cbind(OR = coef(fit), confint(fit)))
```

### Piped water introduction at follow-up in Samama (Mangochi District)

```{r}
# Restrict to Mangochi District follow-up participants with available data
piped_water_data <- part_data %>%
  filter(!(district == "Nsanje" | fu_piped_water_access == "missing")) %>%
  droplevels()

# Define groups
groups <- c("yes", "no")

# Initialize results dataframe
prevalence_summary <- data.frame(
  Piped_Water = groups,
  Positive_Count = NA,
  Total_N = NA,
  Prevalence_Percent = NA,
  CI_Lower = NA,
  CI_Upper = NA
)

# Loop through groups
for (i in seq_along(groups)) {

  grp <- groups[i]

  # Subset to this piped water group
  subset_data <- subset(piped_water_data, fu_piped_water_access == grp)

  # Count positive cases
  pos <- sum(subset_data$fu_multihousehold_strain == "Yes")

  # Total non-missing observations
  total <- sum(!is.na(subset_data$fu_multihousehold_strain))

  # Calculate CI
  test <- prop.test(pos, total)

  prevalence_summary$Positive_Count[i] <- pos
  prevalence_summary$Total_N[i] <- total
  prevalence_summary$Prevalence_Percent[i] <- round(100 * pos / total, 1)
  prevalence_summary$CI_Lower[i] <- round(100 * test$conf.int[1], 1)
  prevalence_summary$CI_Upper[i] <- round(100 * test$conf.int[2], 1)
}

print(prevalence_summary)
```

```{r}
# Chi-squared test
piped_water_tab <- table(
  piped_water_data$fu_multihousehold_strain,
  piped_water_data$fu_piped_water_access
)
piped_water_tab
chisq.test(piped_water_tab)
```

# Isolate data

```{r}
# Import the isolate data
isolate_data_human <- read.csv("data/Malawi_ESBL_E_coli_isolates.csv")

# Label and tidy ESBL isolate data from 2023
isolate_data_human$District <- as.factor(isolate_data_human$District)
isolate_data_human$Year <- as.factor(isolate_data_human$Year)
isolate_data_human$ST <- as.factor(isolate_data_human$ST)
isolate_data_human$popPUNK_cluster <- as.factor(isolate_data_human$popPUNK_cluster)
isolate_data_human$Phylogroup <- as.factor(isolate_data_human$Phylogroup)
isolate_data_human$fu_participant <- factor(isolate_data_human$fu_participant, levels = c(0, 1), labels = c('No', 'Yes'))

# Generate lineage column combining ST and popPUNK cluster
isolate_data_human$Lineage <- paste0("ST", isolate_data_human$ST, "_popPUNK", isolate_data_human$popPUNK_cluster)
isolate_data_human$Lineage <- as.factor(isolate_data_human$Lineage)
```

## Lineages

```{r}
# Print a summary of the lineages
summary(isolate_data_human$Lineage)
```

```{r}
# STs belonging to more than one popPUNK cluster
ST_multiple_popPUNK <- isolate_data_human %>%
  group_by(ST) %>%
  summarise(n_popPUNK = n_distinct(popPUNK_cluster)) %>%
  filter(n_popPUNK > 1)
ST_multiple_popPUNK
```

```{r}
# popPUNK clusters belonging to more than one ST
popPUNK_multiple_ST <- isolate_data_human %>%
  group_by(popPUNK_cluster) %>%
  summarise(n_ST = n_distinct(ST)) %>%
  filter(n_ST > 1)
popPUNK_multiple_ST
```


## Sequence types (MLST)

```{r}
# Plot STs of all isolates
ST_plot <- ggplot(isolate_data_human, aes(x = fct_rev(fct_infreq(ST)), fill = Phylogroup)) +
  geom_bar(color = "black", linewidth = 0.3) +
  coord_flip() +
  theme_bw() +
  labs(x = "ST", y = "Number of isolates") +
  scale_fill_brewer(palette = "Set2")

# Save the plot
ggsave("figures/ST_plot.png", ST_plot, width = 4, height = 5, dpi = 600)

# Print the plot
print(ST_plot)
```

```{r}
# Add year breakdown to ST plot

# Define the number of isolates per year
isolate_totals <- c(`2023` = 76, `2024` = 83)

# Create a dataset of all possible combinations of ST and Year (2023, 2024)
all_st_year_combinations <- expand.grid(ST = unique(isolate_data_human$ST), Year = c(2023, 2024))
all_st_year_combinations$Year <- as.factor(all_st_year_combinations$Year)

# Summarize ST rates per 100 isolates per year
ST_summary_data <- isolate_data_human %>%
  group_by(ST, Year) %>%
  summarise(Count = n(), .groups = "drop") %>%
  full_join(all_st_year_combinations, by = c("ST", "Year")) %>%  # Ensure all ST-Year combinations exist
  mutate(Count = replace_na(Count, 0)) %>%  # Fill missing values with 0
  mutate(rate_per_100 = (Count / isolate_totals[as.character(Year)]) * 100)

# Calculate total counts of each ST to enable plot ordering
ST_summary_data <- ST_summary_data %>%
  group_by(ST) %>%
  mutate(total_count = sum(Count)) %>%
  ungroup()

# Reorder samples in highest to lowest total isolate count order to match the first ST plot
ST_summary_data <- ST_summary_data %>%
  mutate(ST = fct_rev(fct_reorder(ST, total_count, .desc = TRUE)))

# Plot ST rates per 100 participants per year
ST_plot_years <- ggplot(ST_summary_data, aes(x = ST, y = rate_per_100, fill = factor(Year))) +
  geom_bar(color = "black", linewidth = 0.3, stat = "identity", position = position_dodge(width = 0.8)) +
  coord_flip() +
  theme_bw() +
  labs(x = "ST", y = "Percentage of isolates per year", fill = "Year")

# Save the plot
ggsave("figures/ST_plot_years.png", ST_plot_years, width = 4, height = 5, dpi = 600)

# Print the plot
print(ST_plot_years)
```

```{r}
# Fisher exact test with Monte Carlo simulation
st_year_tab <- xtabs(Count ~ ST + Year, data = ST_summary_data)
set.seed(1)
fisher.test(st_year_tab, simulate.p.value = TRUE, B = 1e4)
```

## ARGs

```{r}
# Import and clean ARG data

# Import data
ARGs <- read.csv("data/ARGs.csv", check.names = FALSE)

# Remove '.fasta' from all entries in column 1
ARGs[, 1] <- gsub("\\.fasta", "", ARGs[, 1])

# Apply transformations to all columns except 'Isolate' names so ARGs are present (1) or absent (0)
ARGs[, -1] <- lapply(ARGs[, -1], function(x) {
  ifelse(is.na(x) | x == "." | x == "", 0, 1)
})

# Clean column names
colnames(ARGs) <- gsub("bla", "", colnames(ARGs))  # Remove 'bla'
colnames(ARGs) <- sapply(colnames(ARGs), function(x) {
  paste0(toupper(substring(x, 1, 1)), substring(x, 2))
})  # Capitalize first letter

# Create a base name without suffixes for grouping
base_names <- sub("_.*", "", colnames(ARGs))

# Create a list of indices for columns to merge based on cleaned names
cols_to_merge_list <- split(seq_along(colnames(ARGs)), base_names)

# Track columns to delete after summing
cols_to_delete <- c()

# Sum values into the first column for each group and mark columns for deletion
for (col_indices in cols_to_merge_list) {
  if (length(col_indices) > 1) {
    # Sum values into the first column
    summed_column <- rowSums(ARGs[, col_indices], na.rm = TRUE)
    
    # Set to 1 if total is greater than 1
    summed_column[summed_column > 1] <- 1
    
    # Replace the first occurrence with the summed column
    ARGs[, col_indices[1]] <- summed_column
    
    # Mark all other columns except the first for deletion
    cols_to_delete <- c(cols_to_delete, col_indices[-1])
  }
}

# Delete the marked columns all at once after the loop
ARGs <- ARGs[, -cols_to_delete]

# Clean column names (remove everything after and including '_')
colnames(ARGs) <- gsub("_.*", "", colnames(ARGs))

# Mutate the data to have Isolate and ARG columns
ARGs_long <- ARGs %>%
  pivot_longer(
    cols = -Isolate,
    names_to = "ARG",
    values_to = "Presence"
  )

# Filter for only relevant rows (where presence is 1)
ARGs_long <- ARGs_long %>%
  dplyr::filter(Presence == 1) %>%
  dplyr::select(-Presence)

# Print all unique AMR determinants in data set
(colnames(ARGs)[-1])
```

```{r}
# Define the AMR classes and their associated genes
aminoglycoside = "Aac|Aad|Ant|Aph"
beta_lactam_AmpC = "DHA"
beta_lactam_ESBL = "CTX"
beta_lactam_other = "OXA|TEM"
chloramphenicol = "Cat|FloR"
fosfomycin = "Fos"
macrolide = "Mph"
MDR = "Erm"
quinolone = "Qnr|Qep"
rifampicin = "ARR"
sulphonamide = "Sul"
tetracycline = "Tet"
trimethoprim = "Dfr"
```

```{r}
# Calculate the absolute abundance (counting only where Presence is 1) and label ARG classes
ARG_abundance <- ARGs_long %>%
  group_by(ARG) %>%
  summarise(Count = n(), .groups = "drop") %>%
  mutate(class = case_when(
    str_detect(ARG, aminoglycoside) ~ "Aminoglycoside",
    str_detect(ARG, beta_lactam_AmpC) ~ "Beta-lactam (AmpC)",
    str_detect(ARG, beta_lactam_ESBL) ~ "Beta-lactam (ESBL)",
    str_detect(ARG, beta_lactam_other) ~ "Beta-lactam (other)",
    str_detect(ARG, chloramphenicol) ~ "Chloramphenicol",
    str_detect(ARG, fosfomycin) ~ "Fosfomycin",
    str_detect(ARG, macrolide) ~ "Macrolide",
    str_detect(ARG, MDR) ~ "MDR",
    str_detect(ARG, quinolone) ~ "Quinolone",
    str_detect(ARG, rifampicin) ~ "Rifampicin",
    str_detect(ARG, sulphonamide) ~ "Sulphonamide",
    str_detect(ARG, tetracycline) ~ "Tetracycline",
    str_detect(ARG, trimethoprim) ~ "Trimethoprim",
    TRUE ~ NA_character_
  ))

# Reorder ARGs for plotting: first by class, then by count (highest to lowest)
ARG_abundance <- ARG_abundance %>%
  arrange(class, desc(Count)) %>%
  mutate(ARG = factor(ARG, levels = ARG))

# Reverse the order of classes and ARGs for top-to-bottom plotting
ARG_abundance <- ARG_abundance %>%
  mutate(class = factor(class, levels = sort(unique(class)))) %>%
  mutate(ARG = factor(ARG, levels = rev(ARG[order(ARG_abundance$class, -ARG_abundance$Count)])))

# Create the bar plot
AMR_plot <- ggplot(ARG_abundance, aes(y = ARG, x = Count, fill = class)) +
  geom_bar(color = "black", linewidth = 0.3, stat = "identity") +
  labs(y = "AMR determinant", x = "Number of isolates") +
  theme_bw() +
  guides(fill = guide_legend(title = "Class")) +  
  theme(axis.text.y = element_text(size = 10),
        plot.background = element_rect(fill = "transparent", colour = NA),
        legend.background = element_rect(fill = "transparent", colour = NA)
        ) +
  scale_fill_manual(values=c("#1b9e77", "#fdbf6f", "#7570b3", "#d95f02",
                   "#66a61e", "#e6ab02", "#a6761d", "#e7298a",
                   "#1f78b4", "#b2df8a", "#fb9a99", "#666666",
                   "#cab2d6", "#ffff99")) 

# Save the plot
ggsave("figures/AMR_plot.png", AMR_plot, height = 7, width = 6, dpi = 600)

# Print the plot
print(AMR_plot)
```

```{r}
# Add year breakdown in follow-up participants to ARG plot

# Add year to follow-up ARG data
ARGs <- ARGs %>%
  mutate(Year = case_when(
    grepl("^c", Isolate) ~ 2023,
    grepl("^f", Isolate) ~ 2024,
    TRUE ~ NA_real_
  ))

# Convert to long format
ARGs_fu_long <- ARGs %>%
  pivot_longer(cols = -c(Isolate, Year), names_to = "ARG", values_to = "Presence")

# Summarize counts for each gene per year
ARG_summary_data <- ARGs_fu_long %>%
  group_by(ARG, Year) %>%
  summarize(Count = sum(Presence), .groups = "drop") %>%
  mutate(rate_per_100 = (Count / isolate_totals[as.character(Year)]) * 100) %>%
  mutate(class = case_when(
    str_detect(ARG, aminoglycoside) ~ "Aminoglycoside",
    str_detect(ARG, beta_lactam_AmpC) ~ "Beta-lactam (AmpC)",
    str_detect(ARG, beta_lactam_ESBL) ~ "Beta-lactam (ESBL)",
    str_detect(ARG, beta_lactam_other) ~ "Beta-lactam (other)",
    str_detect(ARG, chloramphenicol) ~ "Chloramphenicol",
    str_detect(ARG, fosfomycin) ~ "Fosfomycin",
    str_detect(ARG, macrolide) ~ "Macrolide",
    str_detect(ARG, MDR) ~ "MDR",
    str_detect(ARG, quinolone) ~ "Quinolone",
    str_detect(ARG, rifampicin) ~ "Rifampicin",
    str_detect(ARG, sulphonamide) ~ "Sulphonamide",
    str_detect(ARG, tetracycline) ~ "Tetracycline",
    str_detect(ARG, trimethoprim) ~ "Trimethoprim",
    TRUE ~ NA_character_
  ))

# Calculate total counts of each ARG to enable plot ordering
ARG_summary_data <- ARG_summary_data %>%
  group_by(ARG) %>%
  mutate(total_count = sum(Count)) %>%
  ungroup() %>%
  filter(total_count > 0)

# Reverse the order of classes and ARGs for top-to-bottom plotting
ARG_summary_data <- ARG_summary_data %>%
  mutate(class = factor(class, levels = sort(unique(class)))) %>%
  group_by(ARG) %>% 
  slice_max(total_count, n = 1) %>%
  ungroup() %>%
  arrange(class, desc(total_count)) %>%
  mutate(ARG = factor(ARG, levels = rev(unique(ARG))))

# Plot ST rates per 100 participants per year
ARG_plot_years <- ggplot(ARG_summary_data, aes(x = ARG, y = rate_per_100, fill = factor(Year))) +
  geom_bar(color = "black", linewidth = 0.3, stat = "identity", position = position_dodge(width = 0.8)) +
  coord_flip() +
  theme_bw() +
  labs(x = "AMR determinant", y = "Percentage of isolates per year", fill = "Year") +
  theme(axis.text.y = element_text(size = 10))  # Adjust y-axis text size for better readability

# Save the plot
ggsave("figures/ARG_plot_years.png", ARG_plot_years, height = 7, width = 6, dpi = 600)

# Print the plot
print(ARG_plot_years)
```

## Plasmid types

```{r}
# Import and clean plasmid type data

# Import data
plasmids <- read.csv("data/Plasmid_data.csv", check.names = FALSE)

# Remove '.fasta' from all entries in column 1
plasmids[, 1] <- gsub("\\.fasta", "", plasmids[, 1])

# Apply transformations to all columns except 'Isolate' names plasmids are present (1) or absent (0)
plasmids[, -1] <- lapply(plasmids[, -1], function(x) {
  ifelse(is.na(x) | x == "." | x == "", 0, 1)
})

# Create a base name without suffixes for grouping
base_names <- sub("_.*", "", colnames(plasmids))

# Create a list of indices for columns to merge based on cleaned names
cols_to_merge_list <- split(seq_along(colnames(plasmids)), base_names)

# Track columns to delete after summing
cols_to_delete <- c()

# Sum values into the first column for each group and mark columns for deletion
for (col_indices in cols_to_merge_list) {
  if (length(col_indices) > 1) {
    # Sum values into the first column
    summed_column <- rowSums(plasmids[, col_indices], na.rm = TRUE)
    
    # Set to 1 if total is greater than 1
    summed_column[summed_column > 1] <- 1
    
    # Replace the first occurrence with the summed column
    plasmids[, col_indices[1]] <- summed_column
    
    # Mark all other columns except the first for deletion
    cols_to_delete <- c(cols_to_delete, col_indices[-1])
  }
}

# Delete the marked columns all at once after the loop
plasmids <- plasmids[, -cols_to_delete]

# Clean column names (remove everything after and including '_')
colnames(plasmids) <- gsub("_.*", "", colnames(plasmids))

# Mutate the data to have Isolate and ARG columns
plasmids_long <- plasmids %>%
  pivot_longer(
    cols = -Isolate,
    names_to = "Plasmid",
    values_to = "Presence"
  )

# filter for only relevant rows (where presence is 1)
plasmids_long <- plasmids_long %>%
  dplyr::filter(Presence == 1) %>%
  dplyr::select(-Presence)

# print all unique AMR determinants in data set
(colnames(plasmids)[-1])
```

```{r}
# Calculate the absolute abundance (counting only where Presence is 1) of plasmids
plasmid_abundance <- plasmids_long %>%
  group_by(Plasmid) %>%                 
  summarise(Count = n(), .groups = "drop")

# Reorder plasmids based on Count
plasmid_abundance <- plasmid_abundance %>%
  mutate(Plasmid = fct_reorder(Plasmid, Count))

# Create the plot
plasmid_plot <- ggplot(plasmid_abundance, aes(y = Plasmid, x = Count)) +
  geom_bar(colour = "black", linewidth = 0.3, stat = "identity") +
  labs(y = "Plasmid replicon", x = "Number of isolates") +
  theme_bw() + 
  theme(axis.text.y = element_text(size = 10))


# Save the plot
ggsave("figures/Plasmid_plot.png", plasmid_plot, height = 5, width = 4, dpi = 600)

# Print the plot
print(plasmid_plot)
```

## Virulence

```{r}
# Import data
virulence <- read.csv("data/virulence.csv", check.names = FALSE)

# Remove '.fasta' from all entries in column 1
virulence[, 1] <- gsub("\\.fasta", "", virulence[, 1])

# Apply transformations to all columns except 'Isolate' names so gene names are present (1) or absent (0)
virulence[, -1] <- lapply(virulence[, -1], function(x) {
  ifelse(is.na(x) | x == "." | x == "", 0, 1)
})

# Inspect data
virulence
```

# Phylogenetic tree with annotation

```{r}
# Read in the treefile
tree <- read.tree("data/human_esbl_core_snps.aln.treefile")

# Root the tree at midpoint
tree <- midpoint(tree)

# Extract isolate metadata for tree
phylogroup_data <- isolate_data_human[, c("Isolate", "Phylogroup")]

# Reorder metadata to match tree tip order
phylogroup_data <- phylogroup_data[match(tree$tip.label, phylogroup_data$Isolate), ]
rownames(phylogroup_data) <- phylogroup_data$Isolate

# Prepare and reorder ESBL gene data for tree plotting
ESBL_one_col <- ARGs %>%
  dplyr::select(Isolate, dplyr::contains("CTX")) %>%
  tidyr::pivot_longer(
    cols = -Isolate,
    names_to = "ESBL_gene",
    values_to = "Presence"
  ) %>%
  dplyr::filter(Presence == 1) %>%
  dplyr::group_by(Isolate) %>%
  dplyr::summarise(
    ESBL_gene = paste(ESBL_gene, collapse = " + "),
    .groups = "drop"
  ) %>%
  dplyr::mutate(layer = "ESBL")

# Extract District data for plotting
district_data <- isolate_data_human %>%
  dplyr::select(Isolate, District) %>%
  dplyr::mutate(layer = "District")

# Extract Year data for plotting
year_data <- isolate_data_human %>%
  dplyr::select(Isolate, Year) %>%
  dplyr::mutate(
    Year = as.factor(Year),
    layer = "Year"
  )

# Generate the basic tree with phylogroup data
p <- ggtree(tree) %<+% phylogroup_data +
  geom_tippoint(
    aes(fill = Phylogroup),
    shape = 21,
    color = "black",
    stroke = 0.25,
    size = 1.8
  ) +
  scale_fill_brewer(name = "Phylogroup", palette = "Set2") +
  geom_treescale(x = 0.02, y = 100, width = 0.01, fontsize = 3, linesize = 0.5)

# Add district
p1 <- p +
  new_scale_fill() +
  geom_fruit(
    data = district_data,
    geom = geom_tile,
    mapping = aes(y = Isolate, x = layer, fill = District),
    color = "black",
    size = 0.15,
    offset = 0.015,
    pwidth = 0.005
  ) +
  scale_fill_brewer(name = "District", palette = "Set1")

# Add year
p2 <- p1 +
  new_scale_fill() +
  geom_fruit(
    data = year_data,
    geom = geom_tile,
    mapping = aes(y = Isolate, x = layer, fill = Year),
    color = "black",
    size = 0.15,
    offset = 0.02,
    pwidth = 0.005
  ) +
  scale_fill_brewer(name = "Year", palette = "Set3")

# Add ESBL genes
p3 <- p2 +
  new_scale_fill() +
  geom_fruit(
    data = ESBL_one_col,
    geom = geom_tile,
    mapping = aes(y = Isolate, x = layer, fill = ESBL_gene),
    color = "black",
    size = 0.15,
    offset = 0.02,
    pwidth = 0.005
  ) +
  scale_fill_brewer(name = "ESBL gene", palette = "Set2")

# Save the plot
ggsave("figures/Tree.png", p3, width = 12, height = 6.5, dpi = 600)

# Print the plot
print(p3)
```


# Mapping lineages between households

```{r}
# Read in household GPS data
households <- read.csv("data/Households_Jittered.csv")

# Read in lineages
interhousehold_lineages <- read.csv("data/Household_clusters.csv")
```

## Samama (Mangochi District) Maps

```{r}
## Create the base map

# Convert Samama household coordinates to sf object and project
samama_sf <- households %>%
  filter(district == "Mangochi") %>%
  st_as_sf(coords = c("GPS_E", "GPS_S"), crs = 4326)

samama_sf_proj <- st_transform(samama_sf, 32736)

# Define and project bounding box
rectangle_coords_samama <- matrix(c(
  35.2166, -14.4150,
  35.2260, -14.4150,
  35.2260, -14.4230,
  35.2166, -14.4230,
  35.2166, -14.4150
), ncol = 2, byrow = TRUE)

rectangle_samama <- st_polygon(list(rectangle_coords_samama)) %>%
  st_sfc(crs = 4326) %>%
  st_transform(32736)

box_samama <- st_bbox(rectangle_samama)
rectangle_samama_proj <- st_as_sfc(box_samama)

# Plot the base map
samama_base_map <- ggplot() +
  geom_sf(data = rectangle_samama_proj, fill = "cornsilk", color = "black") +
  geom_sf(data = samama_sf_proj, color = "black", size = 1.8) +
  annotation_scale(location = "bl", pad_x = unit(0.5, "in"), pad_y = unit(0.4, "in")) +
  annotation_north_arrow(
    location = "bl",
    pad_x = unit(0.4, "in"), pad_y = unit(0.8, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  theme_void()

# Save the base map
ggsave("figures/Samama_map.png", plot = samama_base_map, width = 5, height = 5, dpi = 600)

# Print the base map
print(samama_base_map)
```

### Samama (Mangochi District) lineages

Here, we plot lineages (where groups of isolates are from the same ST and popPUNK cluster) in Mangochi District. Selected lineages must be represented in at least 3 households in either 2023 or 2024.

```{r}
# Prepare the data
samama_clusters <- interhousehold_lineages %>%
  filter(district == "Mangochi")

samama_clusters <- samama_clusters %>%
  left_join(
    households %>% select(household, GPS_S, GPS_E),
    by = "household"
  )

# Convert Samama household coordinates to sf object and project
samama_clusters_sf <- samama_clusters %>%
  st_as_sf(coords = c("GPS_E", "GPS_S"), crs = 4326)

samama_clusters_sf_proj <- st_transform(samama_clusters_sf, 32736)

# Plot the clusters

# Set custom colors for year
year_colors <- c(
  "2023" = "blue",
  "2024" = "red",
  "Both years" = "green3"
)

# Define the clusters to include
clusters_to_plot <- c("ST10_popPUNK5", "ST38_popPUNK20", "ST44_popPUNK223", "ST58_popPUNK11", "ST131_popPUNK4", "ST131_popPUNK25", "ST1015_popPUNK11", "ST2852_popPUNK18", "ST3580_popPUNK18", "ST5614_popPUNK251", "ST8025_popPUNK5", "ST13823_popPUNK1407")

# Create individual plots
cluster_plots <- map(clusters_to_plot, function(clust) {
  cluster_data <- samama_clusters_sf_proj %>%
    filter(cluster == clust)

  ggplot() +
    geom_sf(data = rectangle_samama_proj, fill = "cornsilk", color = "black") +
    geom_sf(data = samama_sf_proj, color = "black", size = 1.5) +
    geom_sf(data = cluster_data, aes(color = year), size = 4.0,
            show.legend = clust == "ST5614_popPUNK251") +  # Show legend only for this cluster
    scale_color_manual(
      values = year_colors,
      breaks = names(year_colors),
      name = "Household colonisation year"
    ) +
    labs(title = paste(clust)) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      legend.position = "bottom",
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14)
    ) +
    guides(color = guide_legend(override.aes = list(size = 4)))
})

# Combine plots into one panel with a single shared legend
samama_panel_plot <- wrap_plots(cluster_plots, ncol = 4, nrow = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Save the panel
ggsave("figures/Samama_lineage_panel_map.png", plot = samama_panel_plot, width = 12, height = 12, dpi = 600)

print(samama_panel_plot)
```

## Mthawira (Nsanje District)

```{r}
## Create the base map

# Convert Mthawira household coordinates to sf object and project
mthawira_sf <- households %>%
  filter(district == "Nsanje") %>%
  st_as_sf(coords = c("GPS_E", "GPS_S"), crs = 4326)

mthawira_sf_proj <- st_transform(mthawira_sf, 32736)

# Define and project bounding box
rectangle_coords_mthawira <- matrix(c(
  35.2850, -16.8440,
  35.3020, -16.8440,
  35.3020, -16.8580,
  35.2850, -16.8580,
  35.2850, -16.8440
), ncol = 2, byrow = TRUE)

rectangle_mthawira <- st_polygon(list(rectangle_coords_mthawira)) %>%
  st_sfc(crs = 4326) %>%
  st_transform(32736)

box_mthawira <- st_bbox(rectangle_mthawira)
rectangle_mthawira_proj <- st_as_sfc(box_mthawira)

# Plot the base map
mthawira_base_map <- ggplot() +
  geom_sf(data = rectangle_mthawira_proj, fill = "cornsilk", color = "black") +
  geom_sf(data = mthawira_sf_proj, color = "black", size = 1.8) +
  annotation_scale(location = "bl", pad_x = unit(0.5, "in"), pad_y = unit(0.4, "in")) +
  annotation_north_arrow(
    location = "bl",
    pad_x = unit(0.4, "in"), pad_y = unit(0.8, "in"),
    style = north_arrow_fancy_orienteering
  ) +
  theme_void()

# Save the base map
ggsave("figures/Mthawira_map.png", plot = mthawira_base_map, width = 5, height = 5, dpi = 600)

# Print the base map
print(mthawira_base_map)
```

### Mthawira (Nsanje District) lineages

Here, we plot lineages (where groups of isolates are from the same ST and popPUNK cluster) in Nsanje District. Selected lineages must be represented in at least 3 households in either 2023 or 2024.

```{r}
# Prepare the data
mthawira_clusters <- interhousehold_lineages %>%
  filter(district == "Nsanje")

mthawira_clusters <- mthawira_clusters %>%
  left_join(
    households %>% select(household, GPS_S, GPS_E),
    by = "household"
  )

# Convert Mthawira household coordinates to sf object and project
mthawira_clusters_sf <- mthawira_clusters %>%
  st_as_sf(coords = c("GPS_E", "GPS_S"), crs = 4326)

mthawira_clusters_sf_proj <- st_transform(mthawira_clusters_sf, 32736)

# Plot the clusters

# Set custom colors for year
year_colors <- c(
  "2023" = "blue",
  "2024" = "red",
  "Both years" = "green3"
)

# Define the clusters to include
clusters_to_plot <- c("ST10_popPUNK92", "ST206_popPUNK41_1328", "ST1015_popPUNK11")

# Create individual plots
cluster_plots <- map(clusters_to_plot, function(clust) {
  cluster_data <- mthawira_clusters_sf_proj %>%
    filter(cluster == clust)

  ggplot() +
    geom_sf(data = rectangle_mthawira_proj, fill = "cornsilk", color = "black") +
    geom_sf(data = mthawira_sf_proj, color = "black", size = 1.5) +
    geom_sf(data = cluster_data, aes(color = year), size = 4.0,
            show.legend = clust == "ST206_popPUNK41_1328") +  # Show legend only for this cluster
    scale_color_manual(
      values = year_colors,
      breaks = names(year_colors),
      name = "Household colonisation year"
    ) +
    labs(title = paste(clust)) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 14),
      legend.position = "bottom",
      legend.text = element_text(size = 14),
      legend.title = element_text(size = 14)
    ) +
    guides(color = guide_legend(override.aes = list(size = 4)))
})

# Combine plots into one panel with a single shared legend
mthawira_panel_plot <- wrap_plots(cluster_plots, ncol = 4, nrow = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Save the panel
ggsave("figures/Mthawira_lineage_panel_map.png", plot = mthawira_panel_plot, width = 12, height = 4, dpi = 600)

print(mthawira_panel_plot)
```

# Strain-level analysis

## Strain identification

```{r}
# Read ANI matrix
ani <- read.csv("data/ANI_matrix.csv", row.names = 1)

# Convert ANI to distance
ani_dist_matrix <- 1 - (as.matrix(ani) / 100)

# Convert to "dist" object
ani_dist_obj <- as.dist(ani_dist_matrix)

# Perform hierarchical clustering (average linkage)
hc <- hclust(ani_dist_obj, method = "average")

# Cut tree at ANI > 99.99% (distance < 0.0001)
clusters <- cutree(hc, h = 0.0001)

# Combine results
strain_groups <- split(names(clusters), clusters)

# Print strain groups
print(strain_groups)
```

```{r}
# Generate a dendrogram of ANI distances

# Open SVG device
svg("figures/ANI_dendrogram.svg", width = 10, height = 12)

# Export dendrogram of ANI hierarchical clustering
plot(hc, main = "ANI Hierarchical Clustering Dendrogram",
     xlab = "Isolate", ylab = "Distance (1 - (ANI/100))", cex = 0.5)
abline(h = 0.0001, col = "red", lty = 2, lwd = 1)

# Close device
dev.off()

# Print plot to R session
plot(hc, main = "ANI Hierarchical Clustering Dendrogram",
     xlab = "Isolate", ylab = "Distance (1 - (ANI/100))", cex = 0.5)
abline(h = 0.0001, col = "red", lty = 2, lwd = 1)
```

```{r}
# Define IDs of strains found in multiple households
multihousehold_strains <- c(2, 3, 4, 5, 8, 9, 10, 12, 13, 14, 15, 18, 19, 24, 26,
                            29, 33, 45, 47, 54, 57, 64, 70, 71, 73, 75, 80)

# Define IDs of strains found 3 or more households
target_strains <- c(4, 5, 10, 12, 14, 18, 24, 26, 29, 45, 57, 64, 70, 73, 75)

# Define strains found in exactly 2 households
two_household_strains <- setdiff(
  multihousehold_strains,
  target_strains
)

# Map isolates to strains
strain_df <- data.frame(
  Isolate = names(clusters),
  strain = clusters
)

# Add strain info and classify
isolate_data_human <- isolate_data_human %>%
  left_join(strain_df, by = "Isolate") %>%
  mutate(
    multihousehold_strain = if_else(
      strain %in% multihousehold_strains,
      "Yes",
      "No"
    )
  ) %>%
  mutate(
    network_strain = if_else(
      strain %in% target_strains,
      "Yes",
      "No"
    )
  )
```

## ExPEC categories per strain

```{r}
# Import ExPEC categories for genes detected
expec_categories <- read.csv("data/ExPEC_categories.csv")

# Get the names of the categories
categories <- unique(expec_categories$Category)

# Remove NA or blank categories
categories <- categories[!is.na(categories) & categories != ""]

# Start new dataframe with isolate names
vf_category_counts <- data.frame(
  Isolate = virulence[[1]]
)

# Count number of genes from each ExPEC category present in each isolate
for(cat in categories){
  
  genes_in_cat <- expec_categories$Gene[
    expec_categories$Category == cat & !is.na(expec_categories$Category)
  ]
  
  genes_present <- intersect(
    genes_in_cat,
    colnames(virulence)
  )
  
  if(length(genes_present) == 0){
    vf_category_counts[[cat]] <- 0
  } else {
    vf_category_counts[[cat]] <- rowSums(
      virulence[, genes_present, drop = FALSE],
      na.rm = TRUE
    )
  }
}

# Calculate total number of ExPEC genes detected per isolate
vf_category_counts$VF_total_count <- rowSums(
  vf_category_counts[, categories, drop = FALSE],
  na.rm = TRUE
)

# Calculate number of ExPEC categories represented per isolate
vf_category_counts$VF_category_count <- rowSums(
  vf_category_counts[, categories, drop = FALSE] > 0,
  na.rm = TRUE
)

# Create isolate-to-strain lookup from ANI clusters
strain_df <- data.frame(
  Isolate = names(clusters),
  strain = clusters
)

# Add strain number to each isolate
vf_category_counts <- vf_category_counts %>%
  left_join(strain_df, by = "Isolate")

# Collapse to strain level 
vf_by_strain_categories <- vf_category_counts %>%
  group_by(strain) %>%
  summarise(
    n_isolates = n(),
    VF_category_count = median(VF_category_count),
    VF_total_count = median(VF_total_count),
    same_VF_category_count = n_distinct(VF_category_count) == 1,
    same_VF_total_count = n_distinct(VF_total_count) == 1,
    across(
      all_of(categories),
      ~ median(.x),
      .names = "{.col}_count"
    ),
    .groups = "drop"
  )

# Check whether total VF counts are conserved within strains
table(vf_by_strain_categories$same_VF_total_count)

# Check whether category counts are conserved within ANI-defined strains
table(vf_by_strain_categories$same_VF_category_count)
```

```{r}
# Add household category (1, 2, or 3+ households)
vf_by_strain_categories <- vf_by_strain_categories %>%
  mutate(
    households = case_when(
      strain %in% target_strains ~ "3_plus",
      strain %in% two_household_strains ~ "2",
      TRUE ~ "1"
    )
  )

# Add category presence/absence columns
vf_by_strain_categories <- vf_by_strain_categories %>%
  mutate(
    across(
      all_of(paste0(categories, "_count")),
      ~ .x > 0,
      .names = "{sub('_count$', '_present', .col)}"
    )
  )
```

### Prevalence across household counts

```{r}
# Calculate prevalence of each ExPEC category across household groups
vf_category_prevalence <- lapply(categories, function(cat){
  
  # Count column for this ExPEC category
  count_col <- paste0(cat, "_count")
  
  # Calculate prevalence of this category in each household group
  summary_stats <- vf_by_strain_categories %>%
    mutate(present = .data[[count_col]] > 0) %>%
    group_by(households) %>%
    summarise(
      n_present = sum(present, na.rm = TRUE),
      n_total = n(),
      prevalence_percent = round(100 * n_present / n_total, 1),
      .groups = "drop"
    )
  
  # Function to format prevalence as n/total (%)
  get_summary <- function(group){
    x <- summary_stats %>% filter(households == group)
    if(nrow(x) == 0){
      return(NA)
    }
    paste0(
      x$n_present, "/", x$n_total,
      " (", x$prevalence_percent, "%)"
    )
  }
  
  # Return summary row for this category
  data.frame(
    Category = cat,
    Household_1 = get_summary("1"),
    Household_2 = get_summary("2"),
    Household_3_plus = get_summary("3_plus")
  )
}) %>%
  bind_rows()

vf_category_prevalence
```

```{r}
# Plot ExPEC category count by household group
ggplot(
  vf_by_strain_categories,
  aes(x = households, y = VF_category_count)
) +
  geom_boxplot() +
  geom_jitter(width = 0.15) +
  labs(
    x = "Household group",
    y = "ExPEC category count"
  ) +
  theme_classic()
```

## Identification of ANI-based medoid isolates for long-read sequencing

```{r}
# Identify medoid isolate from each strain represented in 3+ households

# Create empty results data frame
medoids <- data.frame(strain = integer(),
                      n_isolates = integer(),
                      medoid_isolate = character(),
                      mean_distance = numeric(),
                      stringsAsFactors = FALSE)

for (strain_id in target_strains) {
  # Get isolates belonging to this strain
  members <- names(clusters[clusters == strain_id])
  if (length(members) < 3) next  # skip if less than 3 isolate
  
  # Subset distance matrix for this strain
  submat <- ani_dist_matrix[members, members, drop = FALSE]
  
  # Compute average distance to all other members
  mean_dists <- rowMeans(submat, na.rm = TRUE)
  
  # Identify isolate with smallest mean distance (the medoid)
  medoid <- names(which.min(mean_dists))
  
  # Store result
  medoids <- rbind(medoids, data.frame(
    strain = strain_id,
    n_isolates = length(members),
    medoid_isolate = medoid,
    mean_distance = mean_dists[medoid]
  ))
}

# Print medoid isolates per target strain
print(medoids)
```

## SNP-based medoid comparison (K-12 mapping used for study phylogeny)

```{r}
# Read core SNP matrix
snps <- read.csv("data/core_SNP_matrix.csv", row.names = 1)

# Convert to matrix
snp_dist_matrix <- as.matrix(snps)

# Create empty results data frame
snp_medoids <- data.frame(strain = integer(),
                      n_isolates = integer(),
                      medoid_isolate = character(),
                      mean_distance = numeric(),
                      stringsAsFactors = FALSE)

for (strain_id in target_strains) {
  # Get isolates belonging to this strain
  members <- names(clusters[clusters == strain_id])
  if (length(members) < 3) next  # skip if less than 3 isolate
  
  # Subset distance matrix for this strain
  submat <- snp_dist_matrix[members, members, drop = FALSE]
  
  # Compute average distance to all other members
  mean_dists <- rowMeans(submat, na.rm = TRUE)
  
  # Identify isolate with smallest mean distance (the medoid)
  medoid <- names(which.min(mean_dists))
  
  # Store result
  snp_medoids <- rbind(snp_medoids, data.frame(
    strain = strain_id,
    n_isolates = length(members),
    medoid_isolate = medoid,
    mean_distance = mean_dists[medoid]
  ))
}

# Print medoid isolates per target strain
print(snp_medoids)
```

## Chromosomal SNP distances within strains (hybrid assembly of above medoid isolate used as reference for each strain )

```{r}
# Import within-strain genomic distance data
strain_SNPs <- read.csv("data/Strain_distance.csv", header = TRUE)

# Order strains by maximum pairwise SNPs within each Phylogroup
strain_SNPs$Strain <- factor(
  strain_SNPs$Strain,
  levels = strain_SNPs %>%
    arrange(Phylogroup, desc(Max_pairwise_SNPs_in_strain), Strain) %>%
    pull(Strain) %>%
    unique()
)

# Subset cross-district strains
cross_data <- subset(strain_SNPs, Cross_district_strain == "Yes")

# Dummy data for phylogroup legend
phylo_legend <- data.frame(
  Phylogroup = unique(strain_SNPs$Phylogroup)
)

# Define the number of isolates in each strain
isolate_counts <- data.frame(
  Strain = c("i) ST206", "ii) ST10", "vi) ST10", "xii) ST131", "vii) ST3580",
             "xiv) ST38", "xiii) ST131", "viii) ST1015", "iv) ST44", "xi) ST1015",
             "xv) ST38", "iii) ST8025", "x) ST2852", "ix) ST58", "v) ST13823"),
  Isolates_in_strain = c(5, 6, 3, 5, 4, 6, 3, 4, 4, 3, 15, 3, 4, 7, 3)
)

# Define the number of households each strain was found in
household_counts <- data.frame(
  Strain = c("i) ST206", "ii) ST10", "vi) ST10", "xii) ST131", "vii) ST3580",
             "xiv) ST38", "xiii) ST131", "viii) ST1015", "iv) ST44", "xi) ST1015",
             "xv) ST38", "iii) ST8025", "x) ST2852", "ix) ST58", "v) ST13823"),
  Households = c(5, 4, 3, 5, 3, 6, 3, 4, 4, 3, 13, 3, 3, 7, 3)
)

# Define color for multi-district annotation
arrow_color <- "#E57373"

# Ensure same factor order
isolate_counts$Strain <- factor(isolate_counts$Strain, levels = levels(strain_SNPs$Strain))

# Plot
Strain_SNP_distance_boxplot <- ggplot(strain_SNPs, aes(x = Strain, y = SNP_distance, fill = Phylogroup)) +
  geom_boxplot(width = 0.6, linewidth = 0.3, outlier.size = 0.6, show.legend = FALSE) +
  ylim(0, 49) +  # extended y-limit to make room for circles
  geom_point(data = cross_data, aes(x = Strain, y = 42),
             shape = 25, fill = arrow_color, color = arrow_color, size = 3, show.legend = FALSE) +
  geom_point(data = household_counts, aes(x = Strain, y = 46, size = Households),
             shape = 21, fill = "steelblue", color = "black", show.legend = TRUE) +
  scale_size_continuous(
    name = "No. households",
    range = c(2, 8),
    breaks = c(3, 13)
    ) +
  geom_point(aes(x = NA, y = NA, shape = "Yes"),
             fill = arrow_color, color = arrow_color, size = 3) +
  scale_shape_manual(name = "Cross-district strain", values = c("Yes" = 25)) +
  geom_point(data = phylo_legend, aes(x = NA, y = NA, fill = Phylogroup),
             shape = 22, size = 5, color = "black", inherit.aes = FALSE) +
  scale_fill_brewer(palette = "Set3", name = "Phylogroup") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)) +
  labs(x = "Strain", y = "Genomic distance (no. chromosomal SNPs)")

# Save the plot
ggsave("figures/Strain_SNP_distance_boxplot.png", Strain_SNP_distance_boxplot, height = 4, width = 10, dpi = 600)

# Print the plot
print(Strain_SNP_distance_boxplot)
```

## Strain networks

```{r}
# Build isolate -> Strain mapping from pairwise table
strain_map <- strain_SNPs %>%
  select(Isolate = Isolate1, Strain) %>%
  bind_rows(
    strain_SNPs %>% select(Isolate = Isolate2, Strain)
  ) %>%
  distinct()

# Keep only the required columns for edges
edges <- strain_SNPs %>%
  select(Isolate1, Isolate2, SNP_distance)

# Categorise SNP distances for edge construction
edges <- edges %>%
  mutate(
    category = case_when(
      SNP_distance <= 10 ~ "≤10",
      SNP_distance <= 25 ~ "11–25",
      SNP_distance > 25  ~ "26-50"
    ),
    category = factor(category,
                      levels = c("≤10", "11–25", "26-50"),
                      ordered = TRUE)
  )

# Prepare node metadata and attach Strain
all_isolates <- unique(c(strain_SNPs$Isolate1, strain_SNPs$Isolate2))

nodes <- isolate_data_human %>%
  filter(Isolate %in% all_isolates) %>%
  select(Isolate, District, Year, Participant.ID) %>%
  left_join(strain_map, by = "Isolate")

# Create igraph object
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

# Set seed for reproducible plotting and compute layout coordinate
set.seed(42)
lay <- create_layout(g, layout = "fr")

# Prepare dataframe for label positions just outside each Strain cluster
lay_df <- as.data.frame(lay) %>% filter(!is.na(Strain))

# Compute centroid per strain
centroids <- lay_df %>%
  group_by(Strain) %>%
  summarise(cx = mean(x), cy = mean(y), .groups = "drop")

# Compute cluster radius (max distance from centroid) per strain
spread_df <- lay_df %>%
  left_join(centroids, by = "Strain") %>%
  group_by(Strain, cx, cy) %>%
  summarise(radius = max(sqrt((x - cx)^2 + (y - cy)^2)), .groups = "drop")

# Join and compute label coordinates (1.8 * radius above cluster)
strain_labels <- spread_df %>%
  mutate(
    label_x = cx - radius * 0,
    label_y = cy + radius * 1.8
  )

# Final plot with labels placed above clusters
network_plot <- ggraph(lay) +
  geom_edge_link(aes(colour = category), width = 0.7, alpha = 0.7) +
  geom_node_point(aes(fill = District, colour = Year),
                  shape = 21, size = 2.5, stroke = 1.0) +
  geom_text(
    data = strain_labels,
    aes(x = label_x, y = label_y, label = Strain),
    inherit.aes = FALSE,
    size = 3
  ) +
  scale_edge_color_manual(
    values = c("≤10" = "red",
               "11–25" = "orange",
               "26-50" = "yellow2"),
    breaks = c("≤10", "11–25", "26-50")
  ) +
  scale_fill_manual(
    values = c("Mangochi" = "dodgerblue3", "Nsanje" = "lightblue")
  ) +
  scale_colour_manual(
    values = c("2023" = "black", "2024" = "pink3")
  ) +
  theme_void() +
  labs(edge_color = "No. chromosomal SNPs", fill = "District")

# Save the plot
ggsave("figures/Strain_network_plot.png", network_plot, height = 4, width = 5.5, dpi = 600)

# Print the plot
print(network_plot)
```

## Genomic vs geographic distance within villages

```{r}
# Import within-strain genomic and geographic distance data (for comparisons within same village at the same sampling point)
strain_distance <- read.csv("data/Strain_distance_within_village.csv", header = TRUE)

# Order ST levels by Phylogroup
strain_distance$ST <- factor(
  strain_distance$ST,
  levels = strain_distance %>%
    dplyr::arrange(Phylogroup, ST) %>%
    dplyr::pull(ST) %>%
    unique()
)

# Remove connections between isolates from the same household (n=6)
strain_distance <- strain_distance %>%
  dplyr::filter(Household1 != Household2)

# Print summary of SNP distances
summary(strain_distance$SNP_distance)
```

```{r}
# Calculate Spearman's correlation coefficient and p-value
cor.test(strain_distance$SNP_distance, strain_distance$distance_m, method = "spearman")
```

```{r}
# Add geographic distance groupings
strain_distance <- strain_distance %>%
  mutate(dist_group = case_when(
    distance_m < 200 ~ "<200",
    distance_m >= 200 & distance_m <= 400 ~ "200 - 400",
    distance_m > 400 ~ ">400"
  ))
strain_distance <- strain_distance %>%
  mutate(dist_group = factor(dist_group, levels = c("<200", "200 - 400", ">400")))

# Plot geographic distance vs SNP distance (jittering applied to x axis)
SNP_distance_scatter <- ggplot(strain_distance, aes(x = distance_m, y = SNP_distance, fill = Phylogroup)) +
  geom_jitter(height = 0, width = 5, alpha = 0.65, size = 1.8, shape = 21, colour = "black", stroke = 0.15) +
  scale_fill_brewer(palette = "Set3") +
  scale_x_continuous(name = "Distance between households (metres)") +
  scale_y_continuous(name = "Genomic distance (no. chromosomal SNPs)", limits = c(0, 50)) +
  theme_grey()

# Save the plot
ggsave("figures/SNP_distance_scatter.png", SNP_distance_scatter, height = 4, width = 5, dpi = 600)

# Print the plot
print(SNP_distance_scatter)
```

```{r}
# Check for statistical difference in SNP distance between geographic distance groups
kruskal.test(SNP_distance ~ dist_group, data = strain_distance)
```

```{r}
# Perform post-hoc Dunn's Test with Bonferroni correction for p-values
dunnTest(SNP_distance ~ dist_group, data = strain_distance, method = "bonferroni")
```

```{r}
# Manually create a table from Dunn's test output
pvals <- data.frame(
  group1 = c("<200", "<200", "200 - 400"),
  group2 = c(">400", "200 - 400", ">400"),
  p.adj  = c(0.000407142, 0.677055724, 0.009048447),
  y.position = c(48, 42, 45)
)
pvals$signif <- cut(
  pvals$p.adj,
  breaks = c(-Inf, 0.001, 0.01, 0.05, Inf),
  labels = c("***", "**", "*", "ns")
)

# Plot a violin plot with an overlayed boxplot
SNP_distance_violin <- ggplot(strain_distance, aes(x = dist_group, y = SNP_distance, fill = dist_group)) +
  geom_violin(alpha = 0.6, color = "black", linewidth = 0.2) +
  ylim(0, 50) +
  scale_fill_brewer(type = "seq", palette = 2) +
  labs(x = "Distance between households (metres)", y = "Genomic distance (no. chromosomal SNPs)") +
  theme_minimal() +
  theme(legend.position = "none") +
  stat_pvalue_manual(
    pvals,
    label = "signif",      # shows *, **, ***
    tip.length = 0.02
  )

# Save the plot
ggsave("figures/SNP_distance_violin.png", SNP_distance_violin, height = 4, width = 3.5, dpi = 600)

# Print the plot
print(SNP_distance_violin)
```

## Interhousehold strain maps

```{r}
# Prepare the data (make household variable for isolates from strain strains found in 3+ households)
nodes$household <- sub("\\.[0-9]+$", "", nodes$Participant.ID)
```

### Samama (Mangochi District)

```{r}
# Prepare the data
samama_strains <- nodes %>%
  filter(District == "Mangochi")

samama_strains <- samama_strains %>%
  left_join(
    households %>% select(household, GPS_S, GPS_E),
    by = "household"
  )

# Convert Samama household coordinates to sf object and project
samama_strains_sf <- samama_strains %>%
  st_as_sf(coords = c("GPS_E", "GPS_S"), crs = 4326)

samama_strains_sf_proj <- st_transform(samama_strains_sf, 32736)

# Set custom colors for year
year_colors <- c(
  "2023" = "blue",
  "2024" = "red",
  "Both years" = "green3"
)

# Define the clusters to include
clusters_to_plot <- c("i) ST206", "ii) ST10", "iii) ST8025", "iv) ST44", "v) ST13823",
                      "vi) ST10", "vii) ST3580", "viii) ST1015", "ix) ST58", "x) ST2852",
                      "xii) ST131", "xiii) ST131", "xiv) ST38", "xv) ST38")

# Create individual plots
strain_cluster_plots <- map(clusters_to_plot, function(clust) {
  cluster_data <- samama_strains_sf_proj %>%
    filter(Strain == clust)

  ggplot() +
    geom_sf(data = rectangle_samama_proj, fill = "cornsilk", color = "black") +
    geom_sf(data = samama_sf_proj, color = "black", size = 1.5) +
    geom_sf(data = cluster_data, aes(color = Year), size = 4.0,
            show.legend = clust == "xii) ST131") +  # Show legend only for this cluster
    scale_color_manual(
      values = year_colors,
      breaks = names(year_colors),
      name = "Household colonisation year"
    ) +
    labs(title = paste(clust)) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 18),
      legend.position = "bottom",
      legend.text = element_text(size = 22),
      legend.title = element_text(size = 22)
    ) +
    guides(color = guide_legend(override.aes = list(size = 4)))
})

# Combine plots into one panel with a single shared legend
samama_strain_panel_plot <- wrap_plots(strain_cluster_plots, ncol = 5, nrow = 3) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Save the panel
ggsave("figures/Samama_strain_cluster_panel_map.png", plot = samama_strain_panel_plot, width = 15, height = 12, dpi = 600)

print(samama_strain_panel_plot)
```

### Mthawira (Nsanje District)

```{r}
# Prepare the data
mthawira_strains <- nodes %>%
  filter(District == "Nsanje")

mthawira_strains <- mthawira_strains %>%
  left_join(
    households %>% select(household, GPS_S, GPS_E),
    by = "household"
  )

# Convert mthawira household coordinates to sf object and project
mthawira_strains_sf <- mthawira_strains %>%
  st_as_sf(coords = c("GPS_E", "GPS_S"), crs = 4326)

mthawira_strains_sf_proj <- st_transform(mthawira_strains_sf, 32736)

# Set custom colors for year
year_colors <- c(
  "2023" = "blue",
  "2024" = "red",
  "Both years" = "green3"
)

# Define the clusters to include
clusters_to_plot <- c("i) ST206", "ii) ST10", "viii) ST1015", "xi) ST1015")

# Create individual plots
strain_cluster_plots <- map(clusters_to_plot, function(clust) {
  cluster_data <- mthawira_strains_sf_proj %>%
    filter(Strain == clust)

  ggplot() +
    geom_sf(data = rectangle_mthawira_proj, fill = "cornsilk", color = "black") +
    geom_sf(data = mthawira_sf_proj, color = "black", size = 1.5) +
    geom_sf(data = cluster_data, aes(color = Year), size = 4.0,
            show.legend = FALSE) +  # Show legend only for this cluster) +
    scale_color_manual(
      values = year_colors,
      breaks = names(year_colors),
      name = "Household colonisation year"
    ) +
    labs(title = paste(clust)) +
    theme_void() +
    theme(
      plot.title = element_text(hjust = 0.5, size = 18),
      legend.position = "bottom",
      legend.text = element_text(size = 22),
      legend.title = element_text(size = 22)
    ) +
    guides(color = guide_legend(override.aes = list(size = 4)))
})

# Combine plots into one panel with a single shared legend
mthawira_strain_panel_plot <- wrap_plots(strain_cluster_plots, ncol = 5, nrow = 1) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

# Save the panel
ggsave("figures/Mthawira_strain_cluster_panel_map.png", plot = mthawira_strain_panel_plot, width = 15, height = 4, dpi = 600)

print(mthawira_strain_panel_plot)
```

# ST-specific networks and phylogeny in Malawian context

## ST131

```{r}
# Import ST131 isolate metadata, including isolates from previous studies
ST131_metadata <- read.csv("data/ST131_genomes.csv", header = TRUE, row.names = 1)

# Define the core genome alignment size (columns containing only As, Ts, Cs and Gs)
ST131_cg_size <- 3954219
ST131_recomb_masked_cg_size <- 2834919
```

### Recombination masked core genome SNPs

```{r}
# Load ST131 SNP distance matrix masking recombination sites
ST131_snp_matrix_recomb_masked <- read.csv("data/ST131_malawi_SNP_dists_recombination_masked.csv", header = TRUE, row.names = 1)

# Convert to distance object and cluster
ST131_snp_dist_recomb_masked <- as.dist(ST131_snp_matrix_recomb_masked)
ST131_hc_recomb_masked <- hclust(ST131_snp_dist_recomb_masked, method = "average")

# Identify SNP thresholds representing strict, moderate and loose transmission networks adjusting for recombination-masking
ST131_strict <- ST131_recomb_masked_cg_size*0.000002
ST131_moderate <- ST131_recomb_masked_cg_size*0.000005
ST131_loose <- ST131_recomb_masked_cg_size*0.00002
print(c(ST131_strict, ST131_moderate, ST131_loose))
```

```{r}
# Plot a dendrogram of SNP distances that masks recombination

# Open SVG device
svg("figures/ST131_dendrogram.svg", width = 8, height = 4.5)

# Convert hclust to dendrogram
ST131_dend_recomb_masked <- as.dendrogram(ST131_hc_recomb_masked)

# Generate the dendrogram
plot(ST131_dend_recomb_masked,
     main = "ST131",
     xlab = "",
     ylab = "SNP distance",
     cex = 1,        # leaf labels
     cex.main = 1,   # title
     cex.lab  = 0.8,   # axis labels
     cex.axis = 0.8,   # tick labels
     ylim = c(0, 400))

# Close device
dev.off()

# Print the dendrogram
plot(ST131_dend_recomb_masked,
     main = "ST131",
     xlab = "",
     ylab = "SNP distance",
     cex = 1,        # leaf labels
     cex.main = 1,   # title
     cex.lab  = 0.8,   # axis labels
     cex.axis = 0.8,   # tick labels
     ylim = c(0, 400))
```

```{r}
# Plot an annotated SNP distance heatmap using the above clustering

# Convert SNP matrix to a matrix object for plotting
mat <- as.matrix(ST131_snp_matrix_recomb_masked)

# Typography for annotations and legends
# Legend / annotation TITLE text (bold)
title_gp <- gpar(
  fontsize = 10,
  fontface = "bold",
  fontfamily = "sans"
)

# Legend / annotation LABEL text (normal)
label_gp <- gpar(
  fontsize = 9,
  fontface = "plain",
  fontfamily = "sans"
)

# Prepare the metadata
annotation <- as.data.frame(ST131_metadata)
annotation <- annotation[, c("Donor", "Location", "Study", "Year"), drop = FALSE]
annotation$Year <- as.numeric(annotation$Year)

# Match metadata and matrix orders
annotation <- annotation[colnames(mat), , drop = FALSE]

# Define heatmap colour mappings
snp_color <- colorRamp2(
  c(0, 400),
  c("white", "red")
)

# Generate the range for continuous year annotation
year_color <- colorRamp2(
  range(annotation$Year, na.rm = TRUE),
  c("lightblue", "darkblue")
)

ha_col <- HeatmapAnnotation(
  df = annotation,
  col = list(
    Location = c(
      "Blantyre (urban)"          = "springgreen",
      "Mangochi District (rural)" = "springgreen3",
      "Nsanje District (rural)"   = "green4"
    ),
    Donor = c(
      "Hospital inpatient"        = "tan4",
      "Recent hospital discharge" = "tan3",
      "Community"                 = "tan1"
    ),
    Study = c(
      "This study"   = "gray80",
      "Lewis et al." = "gray50",
      "Musicha et al." = "gray20"
    ),
    Year = year_color
  ),
  annotation_name_gp   = title_gp,
  annotation_legend_param = list(
    Location = list(title_gp = title_gp, labels_gp = label_gp),
    Donor    = list(title_gp = title_gp, labels_gp = label_gp),
    Study    = list(title_gp = title_gp, labels_gp = label_gp),
    Year     = list(title_gp = title_gp, labels_gp = label_gp)
  ),
  gp = gpar(col = "grey30", lwd = 0.3)
)

# Generate the annotated heatmap
ht <- Heatmap(
  mat,
  name = "SNP distance",
  col = snp_color,
  rect_gp = gpar(col = "grey30", lwd = 0.3),
  cluster_rows = ST131_hc_recomb_masked,
  cluster_columns = ST131_hc_recomb_masked,
  top_annotation = ha_col,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  show_row_names    = FALSE,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 4.5),
  column_names_gp = gpar(fontsize = 4.5),
  heatmap_legend_param = list(
    title = "SNPs",
    title_gp  = title_gp,
    labels_gp = label_gp,
    at = c(0, 100, 200, 300, 400)
  )
)

# Save the plot
png(
  filename = "figures/ST131_recombination_masked_SNP_heatmap.png",
  width = 7,
  height = 6,
  units = "in",
  res = 600
)
draw(ht, merge_legend = TRUE)
dev.off()

# Print the plot
draw(ht, merge_legend = TRUE)
```

### Recombination masked networks

```{r}
# Convert matrix to long format
long_snp <- ST131_snp_matrix_recomb_masked %>%
  rownames_to_column(var = "Isolate1") %>%
  pivot_longer(-Isolate1, names_to = "Isolate2", values_to = "SNP_distance")

# Keep only the required columns for edges
edges <- long_snp %>%
  select(Isolate1, Isolate2, SNP_distance)

# Categorise SNP distances for edge construction
edges <- edges %>%
  mutate(
    category = case_when(
      SNP_distance <= ST131_strict ~ "<0.0002% (10 SNPs per 5 Mb)",
      SNP_distance <= ST131_moderate ~ "<0.0005% (25 SNPs per 5 Mb)",
      SNP_distance <= ST131_loose  ~ "<0.002% (100 SNPs per 5 Mb)"
    ),
    category = factor(category,
                      levels = c("<0.0002% (10 SNPs per 5 Mb)",
                                 "<0.0005% (25 SNPs per 5 Mb)",
                                 "<0.002% (100 SNPs per 5 Mb)"),
                      ordered = TRUE)
  ) %>%
  filter(!is.na(category))

# Add weighting to distribute nodes
edges <- edges %>%
  mutate(weight = 1 / (SNP_distance + 1))

# Get node data
nodes <- ST131_metadata %>%
  mutate(
    # Combine Study and isolation year with space + brackets
    `Study (isolation year)` = paste0(Study, " (", Year, ")")
  ) %>%
  rownames_to_column(var = "Isolate")

# Create igraph object
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

# Set seed for reproducible plotting and compute layout coordinate
set.seed(42)
lay <- create_layout(g, layout = "fr")

# Plot networks
network_plot <- ggraph(lay) +
  geom_edge_link(aes(colour = category), width = 0.7, alpha = 0.7) +
  geom_node_point(aes(fill = `Study (isolation year)`, colour = Location),
                  shape = 21, size = 2.5, stroke = 1.0) +
  scale_edge_color_manual(
    values = c("<0.0002% (10 SNPs per 5 Mb)" = "red",
               "<0.0005% (25 SNPs per 5 Mb)" = "orange",
               "<0.002% (100 SNPs per 5 Mb)" = "yellow2"),
    breaks = c("<0.0002% (10 SNPs per 5 Mb)",
               "<0.0005% (25 SNPs per 5 Mb)",
               "<0.002% (100 SNPs per 5 Mb)"
               )
  ) +
  scale_colour_manual(
    values = c("Blantyre (urban)"          = "red",
               "Mangochi District (rural)" = "springgreen",
               "Nsanje District (rural)"   = "green4"
               )
  ) +
  scale_fill_manual(
    values = c('Lewis et al. (2017)' = "grey95",
               'Lewis et al. (2018)' = "grey30",
               'Musicha et al. (2019)' = "pink",
               'Musicha et al. (2020)' = "orange2",
               'This study (2023)' = "skyblue1",
               'This study (2024)' = "blue1"
    )
  ) +
  theme_void() +
  labs(edge_color = "SNPs as % of core genome alignment", color = "Location", fill = "Study (isolation year)")

# Save the plot
ggsave("figures/ST131_network_plot.png", network_plot, height = 4, width = 7, dpi = 600)

# Print the plot
print(network_plot)
```

### Networks without masking recombination

```{r}
# Load ST38 SNP distance matrix including recombination sites
ST131_snp_matrix <- read.csv("data/ST131_malawi_SNP_dists.csv", header = TRUE, row.names = 1)

# Identify SNP thresholds representing strict, moderate and loose transmission networks including recombination sites
ST131_strict_with_recomb <- ST131_cg_size*0.000002
ST131_moderate_with_recomb <- ST131_cg_size*0.000005
ST131_loose_with_recomb <- ST131_cg_size*0.00002
print(c(ST131_strict_with_recomb, ST131_moderate_with_recomb, ST131_loose_with_recomb))
```

```{r}
# Convert matrix to long format
long_snp <- ST131_snp_matrix %>%
  rownames_to_column(var = "Isolate1") %>%
  pivot_longer(-Isolate1, names_to = "Isolate2", values_to = "SNP_distance")

# Keep only the required columns for edges
edges <- long_snp %>%
  select(Isolate1, Isolate2, SNP_distance)

# Categorise SNP distances for edge construction
edges <- edges %>%
  mutate(
    category = case_when(
      SNP_distance <= ST131_strict_with_recomb ~ "<0.0002% (10 SNPs per 5 Mb)",
      SNP_distance <= ST131_moderate_with_recomb ~ "<0.0005% (25 SNPs per 5 Mb)",
      SNP_distance <= ST131_loose_with_recomb  ~ "<0.002% (100 SNPs per 5 Mb)"
    ),
    category = factor(category,
                      levels = c("<0.0002% (10 SNPs per 5 Mb)",
                                 "<0.0005% (25 SNPs per 5 Mb)",
                                 "<0.002% (100 SNPs per 5 Mb)"),
                      ordered = TRUE)
  ) %>%
  filter(!is.na(category))

# Add weighting to distribute nodes
edges <- edges %>%
  mutate(weight = 1 / (SNP_distance + 1))

# Create igraph object
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

# Set seed for reproducible plotting and compute layout coordinate
set.seed(42)
lay <- create_layout(g, layout = "fr")

# Plot networks
network_plot <- ggraph(lay) +
  geom_edge_link(aes(colour = category), width = 0.7, alpha = 0.7) +
  geom_node_point(aes(fill = `Study (isolation year)`, colour = Location),
                  shape = 21, size = 2.5, stroke = 1.0) +
  scale_edge_color_manual(
    values = c("<0.0002% (10 SNPs per 5 Mb)" = "red",
               "<0.0005% (25 SNPs per 5 Mb)" = "orange",
               "<0.002% (100 SNPs per 5 Mb)" = "yellow2"),
    breaks = c("<0.0002% (10 SNPs per 5 Mb)",
               "<0.0005% (25 SNPs per 5 Mb)",
               "<0.002% (100 SNPs per 5 Mb)"
               )
  ) +
  scale_colour_manual(
    values = c("Blantyre (urban)"          = "red",
               "Mangochi District (rural)" = "springgreen",
               "Nsanje District (rural)"   = "green4"
               )
  ) +
  scale_fill_manual(
    values = c('Lewis et al. (2017)' = "grey95",
               'Lewis et al. (2018)' = "grey30",
               'Musicha et al. (2019)' = "pink",
               'Musicha et al. (2020)' = "orange2",
               'This study (2023)' = "skyblue1",
               'This study (2024)' = "blue1"
    )
  ) +
  theme_void() +
  labs(edge_color = "SNPs as % of core genome alignment", color = "Location", fill = "Study (isolation year)")

# Save the plot
ggsave("figures/ST131_network_plot_inc_recomb_sites.png", network_plot, height = 4, width = 7, dpi = 600)

# Print the plot
print(network_plot)
```

### Tree

```{r}
# Read in the treefile
ST131_tree <- read.tree("data/ST131.aln.treefile")

# Root the tree at midpoint
ST131_tree <- midpoint(ST131_tree)

# Extract isolate metadata for tree
tree_metadata <- nodes[, c("Isolate", "Location", "Study (isolation year)")]

# Reorder metadata to match tree tip order
tree_metadata <- tree_metadata[match(ST131_tree$tip.label, tree_metadata$Isolate), ]
rownames(tree_metadata) <- tree_metadata$Isolate

# Add tree tip labels for joining with ggtree
tree_metadata_for_join <- tree_metadata %>%
  tibble::rownames_to_column("label")

# Colour palette 
study_fill_vals <- c(
  'Lewis et al. (2017)'   = "grey95",
  'Lewis et al. (2018)'   = "grey30",
  'Musicha et al. (2019)' = "pink",
  'Musicha et al. (2020)' = "orange2",
  'This study (2023)'     = "skyblue1",
  'This study (2024)'     = "blue1"
)

# Plot tree
p <- ggtree(ST131_tree) %<+% tree_metadata_for_join +
  geom_tippoint(
    aes(
      fill  = `Study (isolation year)`,
      shape = Location
    ),
    colour = "black",
    stroke = 0.3,
    size   = 2.0
  ) +
  scale_fill_manual(values = study_fill_vals, drop = FALSE) +
  scale_shape_manual(
    values = c(
      "Blantyre (urban)"          = 21,
      "Mangochi District (rural)" = 22,
      "Nsanje District (rural)"   = 24
    )
  ) +
  theme_tree2() +
  theme(legend.position = "right") +
  labs(
    fill  = "Study (isolation year)",
    shape = "Location"
  ) +
  geom_treescale(
    x = 0.00005, y = 115, width = 0.001,
    fontsize = 3, linesize = 0.5
  )

# Override shape in key so colours are plotted properly
p <- p +
  guides(
    fill = guide_legend(
      override.aes = list(shape = 21, colour = "black")
    )
  )

# Clean x-axis
p <- p +
  theme(
    axis.line.x  = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

# Add some border space to prevent plot clipping
p <- p + scale_y_continuous(expand = expansion(mult = c(0.03, 0.03)))

# Save the plot
ggsave("figures/ST131_tree.png", p, width = 8, height = 8, dpi = 600)

# Print the plot
print(p)
```

## ST38

```{r}
# Import ST38 isolate metadata, including isolates from previous studies
ST38_metadata <- read.csv("data/ST38_genomes.csv", header = TRUE, row.names = 1)

# Define the core genome alignment size (columns containing only As, Ts, Cs and Gs)
ST38_cg_size <- 4279355
ST38_recomb_masked_cg_size <- 3472800
```

### Recombination masked core genome SNPs

```{r}
# Load ST38 SNP distance matrix masking recombination sites
ST38_snp_matrix_recomb_masked <- read.csv("data/ST38_malawi_SNP_dists_recombination_masked.csv", header = TRUE, row.names = 1)

# Convert to distance object and cluster
ST38_snp_dist_recomb_masked <- as.dist(ST38_snp_matrix_recomb_masked)
ST38_hc_recomb_masked <- hclust(ST38_snp_dist_recomb_masked, method = "average")

# Identify SNP thresholds representing strict, moderate and loose transmission networks adjusting for recombination-masking
ST38_strict <- ST38_recomb_masked_cg_size*0.000002
ST38_moderate <- ST38_recomb_masked_cg_size*0.000005
ST38_loose <- ST38_recomb_masked_cg_size*0.00002
print(c(ST38_strict, ST38_moderate, ST38_loose))
```

```{r}
# Plot a dendrogram of SNP distances that masks recombination

# Open SVG device
svg("figures/ST38_dendrogram.svg", width = 8, height = 4.5)

# Convert hclust to dendrogram
ST38_dend_recomb_masked <- as.dendrogram(ST38_hc_recomb_masked)

# Generate the dendrogram
plot(ST38_dend_recomb_masked,
     main = "ST38",
     xlab = "",
     ylab = "SNP distance",
     cex = 1,        # leaf labels
     cex.main = 1,   # title
     cex.lab  = 0.8,   # axis labels
     cex.axis = 0.8,   # tick labels
     ylim = c(0, 400))

# Close device
dev.off()

# Plot the dendrogram
plot(ST38_dend_recomb_masked,
     main = "ST38",
     xlab = "",
     ylab = "SNP distance",
     cex = 1,        # leaf labels
     cex.main = 1,   # title
     cex.lab  = 0.8,   # axis labels
     cex.axis = 0.8,   # tick labels
     ylim = c(0, 400))
```

```{r}
# Plot an annotated SNP distance heatmap using the above clustering

# Convert SNP matrix to a matrix object for plotting
mat <- as.matrix(ST38_snp_matrix_recomb_masked)

# Typography for annotations and legends
# Legend / annotation TITLE text (bold)
title_gp <- gpar(
  fontsize = 10,
  fontface = "bold",
  fontfamily = "sans"
)

# Legend / annotation LABEL text (normal)
label_gp <- gpar(
  fontsize = 9,
  fontface = "plain",
  fontfamily = "sans"
)

# Prepare the metadata
annotation <- as.data.frame(ST38_metadata)
annotation <- annotation[, c("Donor", "Location", "Study", "Year"), drop = FALSE]
annotation$Year <- as.numeric(annotation$Year)

# Match metadata and matrix orders
annotation <- annotation[colnames(mat), , drop = FALSE]

# Define heatmap colour mappings
snp_color <- colorRamp2(
  c(0, 400),
  c("white", "red")
)

# Generate the range for continuous year annotation
year_color <- colorRamp2(
  range(annotation$Year, na.rm = TRUE),
  c("lightblue", "darkblue")
)

ha_col <- HeatmapAnnotation(
  df = annotation,
  col = list(
    Location = c(
      "Blantyre (urban)"          = "springgreen",
      "Mangochi District (rural)" = "springgreen3",
      "Nsanje District (rural)"   = "green4"
    ),
    Donor = c(
      "Hospital inpatient"        = "tan4",
      "Recent hospital discharge" = "tan3",
      "Community"                 = "tan1"
    ),
    Study = c(
      "This study"   = "gray80",
      "Lewis et al." = "gray50",
      "Musicha et al." = "gray20"
    ),
    Year = year_color
  ),
  annotation_name_gp   = title_gp,
  annotation_legend_param = list(
    Location = list(title_gp = title_gp, labels_gp = label_gp),
    Donor    = list(title_gp = title_gp, labels_gp = label_gp),
    Study    = list(title_gp = title_gp, labels_gp = label_gp),
    Year     = list(title_gp = title_gp, labels_gp = label_gp)
  ),
  gp = gpar(col = "grey30", lwd = 0.3)
)


# Generate the annotated heatmap
ht <- Heatmap(
  mat,
  name = "SNP distance",
  col = snp_color,
  rect_gp = gpar(col = "grey30", lwd = 0.3),
  cluster_rows = ST38_hc_recomb_masked,
  cluster_columns = ST38_hc_recomb_masked,
  top_annotation = ha_col,
  show_row_dend = FALSE,
  show_column_dend = FALSE,
  show_row_names    = FALSE,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 4.5),
  column_names_gp = gpar(fontsize = 4.5),
  heatmap_legend_param = list(
    title = "SNPs",
    title_gp  = title_gp,
    labels_gp = label_gp,
    at = c(0, 100, 200, 300, 400)
  )
)

# Save the plot
png(
  filename = "figures/ST38_recombination_masked_SNP_heatmap.png",
  width = 7,
  height = 6,
  units = "in",
  res = 600
)
draw(ht, merge_legend = TRUE)
dev.off()

# Print the plot
draw(ht, merge_legend = TRUE)
```

### Recombination masked networks

```{r}
# Convert matrix to long format
long_snp <- ST38_snp_matrix_recomb_masked %>%
  rownames_to_column(var = "Isolate1") %>%
  pivot_longer(-Isolate1, names_to = "Isolate2", values_to = "SNP_distance")

# Keep only the required columns for edges
edges <- long_snp %>%
  select(Isolate1, Isolate2, SNP_distance)

# Categorise SNP distances for edge construction
edges <- edges %>%
  mutate(
    category = case_when(
      SNP_distance <= ST38_strict ~ "<0.0002% (10 SNPs per 5 Mb)",
      SNP_distance <= ST38_moderate ~ "<0.0005% (25 SNPs per 5 Mb)",
      SNP_distance <= ST38_loose  ~ "<0.002% (100 SNPs per 5 Mb)"
    ),
    category = factor(category,
                      levels = c("<0.0002% (10 SNPs per 5 Mb)",
                                 "<0.0005% (25 SNPs per 5 Mb)",
                                 "<0.002% (100 SNPs per 5 Mb)"),
                      ordered = TRUE)
  ) %>%
  filter(!is.na(category))

# Add weighting to distribute nodes
edges <- edges %>%
  mutate(weight = 1 / (SNP_distance + 1))

# Get node data
nodes <- ST38_metadata %>%
  mutate(
    # Combine Study and isolation year with space + brackets
    `Study (isolation year)` = paste0(Study, " (", Year, ")")
  ) %>%
  rownames_to_column(var = "Isolate")

# Create igraph object
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

# Set seed for reproducible plotting and compute layout coordinate
set.seed(42)
lay <- create_layout(g, layout = "fr")

# Plot networks
network_plot <- ggraph(lay) +
  geom_edge_link(aes(colour = category), width = 0.7, alpha = 0.7) +
  geom_node_point(aes(fill = `Study (isolation year)`, colour = Location),
                  shape = 21, size = 2.5, stroke = 1.0) +
  scale_edge_color_manual(
    values = c("<0.0002% (10 SNPs per 5 Mb)" = "red",
               "<0.0005% (25 SNPs per 5 Mb)" = "orange",
               "<0.002% (100 SNPs per 5 Mb)" = "yellow2"),
    breaks = c("<0.0002% (10 SNPs per 5 Mb)",
               "<0.0005% (25 SNPs per 5 Mb)",
               "<0.002% (100 SNPs per 5 Mb)"
               )
  ) +
  scale_colour_manual(
    values = c("Blantyre (urban)"          = "red",
               "Mangochi District (rural)" = "springgreen",
               "Nsanje District (rural)"   = "green4"
               )
  ) +
  scale_fill_manual(
    values = c('Lewis et al. (2017)' = "grey95",
               'Lewis et al. (2018)' = "grey30",
               'Musicha et al. (2019)' = "pink",
               'Musicha et al. (2020)' = "orange2",
               'This study (2023)' = "skyblue1",
               'This study (2024)' = "blue1"
    )
  ) +
  theme_void() +
  labs(edge_color = "SNPs as % of core genome alignment", color = "Location", fill = "Study (isolation year)")

# Save the plot
ggsave("figures/ST38_network_plot.png", network_plot, height = 4, width = 7, dpi = 600)

# Print the plot
print(network_plot)
```

### Networks without masking recombination

```{r}
# Load ST38 SNP distance matrix including recombination sites
ST38_snp_matrix <- read.csv("data/ST38_malawi_SNP_dists.csv", header = TRUE, row.names = 1)

# Identify SNP thresholds representing strict, moderate and loose transmission networks including recombination sites
ST38_strict_with_recomb <- ST38_cg_size*0.000002
ST38_moderate_with_recomb <- ST38_cg_size*0.000005
ST38_loose_with_recomb <- ST38_cg_size*0.00002
print(c(ST38_strict_with_recomb, ST38_moderate_with_recomb, ST38_loose_with_recomb))
```

```{r}
# Convert matrix to long format
long_snp <- ST38_snp_matrix %>%
  rownames_to_column(var = "Isolate1") %>%
  pivot_longer(-Isolate1, names_to = "Isolate2", values_to = "SNP_distance")

# Keep only the required columns for edges
edges <- long_snp %>%
  select(Isolate1, Isolate2, SNP_distance)

# Categorise SNP distances for edge construction
edges <- edges %>%
  mutate(
    category = case_when(
      SNP_distance <= ST38_strict_with_recomb ~ "<0.0002% (10 SNPs per 5 Mb)",
      SNP_distance <= ST38_moderate_with_recomb ~ "<0.0005% (25 SNPs per 5 Mb)",
      SNP_distance <= ST38_loose_with_recomb  ~ "<0.002% (100 SNPs per 5 Mb)"
    ),
    category = factor(category,
                      levels = c("<0.0002% (10 SNPs per 5 Mb)",
                                 "<0.0005% (25 SNPs per 5 Mb)",
                                 "<0.002% (100 SNPs per 5 Mb)"),
                      ordered = TRUE)
  ) %>%
  filter(!is.na(category))

# Add weighting to distribute nodes
edges <- edges %>%
  mutate(weight = 1 / (SNP_distance + 1))

# Create igraph object
g <- graph_from_data_frame(d = edges, vertices = nodes, directed = FALSE)

# Set seed for reproducible plotting and compute layout coordinate
set.seed(42)
lay <- create_layout(g, layout = "fr")

# Plot networks
network_plot <- ggraph(lay) +
  geom_edge_link(aes(colour = category), width = 0.7, alpha = 0.7) +
  geom_node_point(aes(fill = `Study (isolation year)`, colour = Location),
                  shape = 21, size = 2.5, stroke = 1.0) +
  scale_edge_color_manual(
    values = c("<0.0002% (10 SNPs per 5 Mb)" = "red",
               "<0.0005% (25 SNPs per 5 Mb)" = "orange",
               "<0.002% (100 SNPs per 5 Mb)" = "yellow2"),
    breaks = c("<0.0002% (10 SNPs per 5 Mb)",
               "<0.0005% (25 SNPs per 5 Mb)",
               "<0.002% (100 SNPs per 5 Mb)"
               )
  ) +
  scale_colour_manual(
    values = c("Blantyre (urban)"          = "red",
               "Mangochi District (rural)" = "springgreen",
               "Nsanje District (rural)"   = "green4"
               )
  ) +
  scale_fill_manual(
    values = c('Lewis et al. (2017)' = "grey95",
               'Lewis et al. (2018)' = "grey30",
               'Musicha et al. (2019)' = "pink",
               'Musicha et al. (2020)' = "orange2",
               'This study (2023)' = "skyblue1",
               'This study (2024)' = "blue1"
    )
  ) +
  theme_void() +
  labs(edge_color = "SNPs as % of core genome alignment", color = "Location", fill = "Study (isolation year)")

# Save the plot
ggsave("figures/ST38_network_plot_inc_recomb_sites.png", network_plot, height = 4, width = 7, dpi = 600)

# Print the plot
print(network_plot)
```

### Tree

```{r}
# Read in the treefile
ST38_tree <- read.tree("data/ST38.aln.treefile")

# Root the tree at midpoint
ST38_tree <- midpoint(ST38_tree)

# Extract isolate metadata for tree
tree_metadata <- nodes[, c("Isolate", "Location", "Study (isolation year)")]

# Reorder metadata to match tree tip order
tree_metadata <- tree_metadata[match(ST38_tree$tip.label, tree_metadata$Isolate), ]
rownames(tree_metadata) <- tree_metadata$Isolate

# Add tree tip labels for joining with ggtree
tree_metadata_for_join <- tree_metadata %>%
  tibble::rownames_to_column("label")

# Colour palette 
study_fill_vals <- c(
  'Lewis et al. (2017)'   = "grey95",
  'Lewis et al. (2018)'   = "grey30",
  'Musicha et al. (2019)' = "pink",
  'Musicha et al. (2020)' = "orange2",
  'This study (2023)'     = "skyblue1",
  'This study (2024)'     = "blue1"
)

# Plot tree
p <- ggtree(ST38_tree) %<+% tree_metadata_for_join +
  geom_tippoint(
    aes(
      fill  = `Study (isolation year)`,
      shape = Location
    ),
    colour = "black",
    stroke = 0.3,
    size   = 2.0
  ) +
  scale_fill_manual(values = study_fill_vals, drop = FALSE) +
  scale_shape_manual(
    values = c(
      "Blantyre (urban)"          = 21,
      "Mangochi District (rural)" = 22,
      "Nsanje District (rural)"   = 24
    )
  ) +
  theme_tree2() +
  theme(legend.position = "right") +
  labs(
    fill  = "Study (isolation year)",
    shape = "Location"
  ) +
  geom_treescale(
    x = 0.0005, y = 50, width = 0.001,
    fontsize = 3, linesize = 0.5
  )

# Override shape in key so colours are plotted properly
p <- p +
  guides(
    fill = guide_legend(
      override.aes = list(shape = 21, colour = "black")
    )
  )

# Clean x-axis
p <- p +
  theme(
    axis.line.x  = element_blank(),
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  )

# Add some border space to prevent plot clipping
p <- p + scale_y_continuous(expand = expansion(mult = c(0.03, 0.03)))

# Save the plot
ggsave("figures/ST38_tree.png", p, width = 8, height = 8, dpi = 600)

# Print the plot
print(p)
```
